Modelling last glacial cycle ice dynamics in the Alps

### 2017-07-18 articles/cycle/alps-cycle.tex
Add tentative introduction, a big mess.

% ----------------------------------------------------------------------
\section{Introduction}
% ----------------------------------------------------------------------

    \begin{itemize}
    \item The European Alps are the cradle of pioneer glacial studies.
    \item Today, their glacial history is better understood than any other.
    \item The LGM extent is very well known (Fig.~\ref{fig:lgmvel}a).
    \item But modelling was difficult due to the steep topography.
    \item Here, we us PISM to model the last glacial cycle (120--0\,ka).
    \end{itemize}

@@

% ----------------------------------------------------------------------
\section{Ice sheet model set-up}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Palaeo-climate forcing}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Glacier dynamics}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Erosion potential}
% ----------------------------------------------------------------------

>>>>

% ----------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ----------------------------------------------------------------------

    Glaciers move by a combination of meltwater-induced sliding at the base
    \citep[\S532]{Saussure.1779}, and viscous deformation within the ice body
    \citep{Forbes.1846b}. As glaciers flow and slide across their bed, they
    transport rock debris and erode the landscape, thereby leaving
    geomorphological traces of their former presence.

    For nearly 300~years, montane people and early explorers of the European
    Alps learned to read the geomorphological imprint left by glaciers in the
    landscape, and to understand that glaciers had once been more extensive
    than today \citep[e.g.,][p.~21]{Windham.Martel.1744}. In the mid-nineteenth
    century, more systematic studies of glacial features showed that alpine
    glaciers extended well below their contemporary margins \citep{Venetz.1821}
    and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
    idea that, under colder temperatures, expansive ice sheets had once covered
    much of Europe and North America \citep{Agassiz.1840}.

    However, this glacial theory did not gain general acceptance until the
    discovery and exploration of the two present-day ice sheets on Earth, the
    Greenland and Antarctic ice sheets, provided a modern analogue for the
    proposed European and North American ice sheets. Although it was long
    unclear whether there had been a single or multiple glaciations, this
    controversy ended with the large-scale mapping of two distinct moraine
    systems in North America \citep{Chamberlin.1894}. In the European Alps, the
    systematic classification of the glaciofluvial terraces on the northern
    foreland indicated that landscape, that there had been at least four major
    glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
    of glaciofluvial stratigraphy in the foreland indicate up to eight
    glaciations \citep{Ivy-Ochs.etal.2008, Preusser.etal.2011}.

    More recently, palaeoclimate records extracted from deep sea sediments and
    ice cores have provided a much more detailed picture of the Earth
    environmental history \citep[e.g.,][]{Emiliani.1955,
    Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
    indicating neither one, nor four, but rather tens of glacial cycles. During
    the last 800000 years (800\,ka), glacial and interglacial periods have
    succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
    Augustin.etal.2004}. However, much of the traces left by glaciers on the
    landscape date from the most recent of these cycles, the \emph{last glacial
    cycle}.

    Nevertheless, the landform record is sparse in time and, most often,
    spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
    on the lanscape, and much of this evidence has been overprinted by
    subsequent glacier re-advances and other geomorphological processes
    \citep{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. The landscape
    imprint of previous glaciations has been largely overprinted by the more
    recent ones; therefore most of the glacial features currently left on the
    foreland present a record of the last major glaciation of the Alps, dating
    from the Last Glacial Maximum (LGM).

    Because this global signal is largely governed by the North American and
    Eurasian ice sheet complexes, it is unclear wether glacier advances and
    retreats in the Alps were in pac(71--57\,ka)e with global sea-level
    fluctuations. The timing of the LGM in the Alps (Ivy-Ochs et al., 2008;
    Monegato et al., 2007) is in good agreement with the maximum expansion of
    continental ice sheets recorded by the Marine Oxygen Isotope Stage (MIS) 2
    between 29 000 and 14 000 years before present (BP) (Lisiecki and Raymo,
    2005). There are nonetheless regional variations between different piedmont
    lobes, that are of the same order of magnitude with the dating
    uncertainties. Because recent glaciation overprinted older ones, and
    because dating uncertainties typically increase with age, dated
    reconstructions are strongly biased towards earlier glaciations
    \citep{Heyman.etal.2011}, and thus much of the glacial records dates to the
    Last Glacial Maximum (LGM). The Alps have been studied in more detail than
    any other glacial region. There is evidence for glaciation during MIS 4
    (71--57\,ka) but potential foreland glaciations during the rest of the last
    glacial cycle are still discussed.

    Here, we use the Parallel Ice Sheet Model
    \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
    approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
    model alpine glacier dynamics through the last glacial cycle (120--0\,ka),
    a period for which palaeo-temperature proxies are available, albeit non
    regional. We test the model sensitivity to multiple palaeo-climate forcing
    (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
    at high resolution for the optimal forcing (Sect.~\ref{sec:results}), and
    the resulting glacial erosion pattern (Sect.~\ref{sec:erosion}).

@@

% ----------------------------------------------------------------------
\section{Ice sheet model set-up}
\label{sec:model}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Palaeo-climate forcing}
\label{sec:climate}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Glacier dynamics}
\label{sec:results}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Erosion potential}
\label{sec:erosion}
% ----------------------------------------------------------------------
